### 2025-02-08 tfp2025 3/src/sections/1-introduction.tex
work

\begin{code}[hide]%
\>[0]\AgdaSymbol{\{-\#}\AgdaSpace{}%
\AgdaKeyword{OPTIONS}\AgdaSpace{}%
\AgdaPragma{--cubical}\AgdaSpace{}%
\AgdaPragma{--guarded}\AgdaSpace{}%
\AgdaSymbol{\#-\}}\<%
\\
\>[0]\AgdaKeyword{module}\AgdaSpace{}%
\AgdaModule{1-introduction}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[0]\<%
\end{code}

@@

Parsing is the conversion of flat, human-readable text into a tree structure
that is easier for computers to manipulate.  As one of the central
pillars of compiler tooling since the 1960s, today almost every automated
transformation of computer programs requires a form of parsing.
Though it is such a mature research subject, it is still actively studied, for example the question of how to resolve ambiguities in context-free grammars \cite{one-parser-to-rule-them-all}.

Recent work by Elliot uses interactive theorem provers to state simple specifications of languages and that proofs of desirable properties of these language specifications transfer easily to their parsers \cite{conal-languages}. Unfortunately, this work only considers regular languages which are not powerful enough to describe practical programming languages.

In this paper, we formalize context-free languages and show how to parse them, extending Elliot’s type theoretic approach to language specfication.  One of the main challenges is that the recursive nature of context-free languages does not map directly onto automated theorem provers as they do not support general recursion. We use a fuel-based approach to solve this problem.

@@

For this paper we have chosen Agda as our type theory and interactive theorem prover. We believe our definitions should transfer easily to other theories and tools. This paper itself is a literate Agda file; all highlighted Agda code has been accepted by Agda's type checker, giving us a high confidence of correctness. Unfortunately, we are still working out the proof of three postulates in \cref{sec:cfg-parsing}. These are the only postulates that we have yet to prove.

>>>>

\begin{code}[hide]%
\>[0]\AgdaSymbol{\{-\#}\AgdaSpace{}%
\AgdaKeyword{OPTIONS}\AgdaSpace{}%
\AgdaPragma{--cubical}\AgdaSpace{}%
\AgdaPragma{--guarded}\AgdaSpace{}%
\AgdaSymbol{\#-\}}\<%
\\
\>[0]\AgdaKeyword{module}\AgdaSpace{}%
\AgdaModule{1-introduction}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
%
\\[\AgdaEmptyExtraSkip]%
\>[0]\AgdaKeyword{open}\AgdaSpace{}%
\AgdaKeyword{import}\AgdaSpace{}%
\AgdaModule{Data.Unit}\<%
\\
\>[0]\<%
\end{code}

@@

Parsing is the conversion of flat, human-readable text into a tree structure
that is easier for computers to manipulate.  As one of the central
pillars of compiler tooling since the 1960s, today almost every automated
transformation of computer programs requires a form of parsing.
Though it is a mature research subject, it is still actively studied, for example the question of how to resolve ambiguities in context-free grammars \cite{one-parser-to-rule-them-all}.

Most parsing works mix the essence of the parsing technique with operational details \jr{such as... state machines, continuations, memoization?}. Our understanding and ability to improve upon these parsing techniques is hindered by the additional complexity of these inessential practical concerns. To address this issue, we are developing natural denotational semantics for traditional parsing techniques.

\jr{Elliot has kicked off this effort...}
Recent work by Elliot uses interactive theorem provers to state simple specifications of languages and that proofs of desirable properties of these language specifications transfer easily to their parsers \cite{conal-languages}. Unfortunately, this work only considers regular languages which are not powerful enough to describe practical programming languages.

\jr{Make the problem clear through an example: if we have a left-recursive grammar then naively unfolding it gets us into an infinite loop. }

In this paper, we formalize context-free languages and show how to parse them, extending Elliot’s type theoretic approach to language specfication.  One of the main challenges is that the recursive nature of context-free languages does not map directly onto interactive theorem provers as they do not support general recursion (for good reasons). We encode context-free languages as fixed points of functors (initial algebras).

\jr{Say something about the limitation that we only study acyclic grammars: there must be a total order on nonterminals and a nonterminal is not allowed to refer to nonterminals that come before it. We wanted to start by limiting ourselves to grammars with only one nonterminal, but those are not closed under derivatives.}

@@

For this paper we have chosen Agda as our type theory and interactive theorem prover. We believe our definitions should transfer easily to other theories and tools. This paper itself is a literate Agda file; all highlighted Agda code has been accepted by Agda's type checker, giving us a high confidence of correctness.

% The goal is to give a denotational semantics to context-free languages
% And mechanize this in a proof assistants

% Challenge:
% Give a simple non-recursive example
% Expand to a recursive variant

% We could use full blown domain theory, but that is quite a big hammer

% In section 2:
% * Could: "This is a background section"
% * We recall Elliot's ...
% * To make it easier to add fixed points in the next section

% Look for a simple example in Section 2. Can be contrived

### 2025-02-08 tfp2025 3/src/sections/4-discussion.tex
work

\paragraph{Expressiveness} We conjecture that our grammars which include variables and fixed points can describe any context-free language. We have shown the example of balanced the bracket language which is known to be context-free. Furthermore, Grenrus shows that any context-free grammar can be converted to his grammars \cite{fix-ing-regular-expressions}, which are similar to our grammars. The main problem is showing that mutually recursive nonterminals can be expressed using our simple fixed points, which requires Beki\'c's bisection lemma~\cite{Bekic1984}. Formalizing this in our framework is future work.

@@

\paragraph{Performance}
For a parser to practically useful, it must at least have linear asymptotic complexity for practical grammars. Might et al. \cite{parsing-with-derivatives} show that naively parsing using derivatives does not achieve that bound, but optimizations might make it possible. In particular, they argue that we could achieve $O(n|G|)$ time complexity (where $|G|$ is the grammar size) if the grammar size stays approximately constant after every derivative. By compacting the grammar, they conjecture it is possible to achieve this bound for any unambiguous grammar. We want to investigate if similar optimizations could be applied to our parser and if we can prove that we achieve this bound.

\paragraph{Simplicity}
One of the main contributions of Elliot's type theoretic formalization of languages~\cite{conal-languages} is its simplicity of implementation and proof. To be able to extend his approach to context-free languages we have had to introduce some complications. Most notably, we use fuel to define the semantics of our grammars. We have explored other approaches such as using guarded type theory, but we did not manage to significantly simplify our formalization. Furtheremore, we expect that many proofs remain simple despite our fuel-based approach.
\medbreak

In conclusion, we have (almost) formalized context-free grammars using a type theoretic approach to provide fertile ground for further formalizations of disambiguation strategies and parsers that are both correct and performant.

>>>>

\jr{TODO: μ-regular expressions have been studied before, cite}
\paragraph{Expressiveness} We conjecture that our grammars which include variables and fixed points can describe any context-free language. We have shown the example of balanced the bracket language which is known to be context-free. Furthermore, Grenrus shows that any context-free grammar can be converted to his grammars \cite{fix-ing-regular-expressions}, which are similar to our grammars. The main problem is showing that mutually recursive nonterminals can be expressed using our simple fixed points, which requires Beki\'c's bisection lemma~\cite{Bekic1984}. Formalizing this in our framework is future work.

@@

\paragraph{Performance}
\jr{cite Jeremy Yallop's work}
For a parser to practically useful, it must at least have linear asymptotic complexity for practical grammars. Might et al. \cite{parsing-with-derivatives} show that naively parsing using derivatives does not achieve that bound, but optimizations might make it possible. In particular, they argue that we could achieve $O(n|G|)$ time complexity (where $|G|$ is the grammar size) if the grammar size stays approximately constant after every derivative. By compacting the grammar, they conjecture it is possible to achieve this bound for any unambiguous grammar. We want to investigate if similar optimizations could be applied to our parser and if we can prove that we achieve this bound.

\paragraph{Simplicity}
One of the main contributions of Elliot's type theoretic formalization of languages~\cite{conal-languages} is its simplicity of implementation and proof. To be able to extend his approach to context-free languages we have had to introduce some complications.
\jr{TODO: finish this paragraph}
% Most notably, we use fuel to define the semantics of our grammars. We have explored other approaches such as using guarded type theory, but we did not manage to significantly simplify our formalization. Furtheremore, we expect that many proofs remain simple despite our fuel-based approach.
\medbreak

In conclusion, we have formalized (acyclic) context-free grammars using a type theoretic approach to provide fertile ground for further formalizations of disambiguation strategies and parsers that are both correct and performant.
